PineAPPL: combining EW and QCD corrections\\ for fast evaluation of LHC processes

### 2020-08-28 paper/sec-pineappl.tex
Add section reference
Fix missing reference
Add last-minute changes to section 2

In this section we introduce \textsc{PineAPPL}.
We first describe the general scope and features of the library in comparison to \textsc{APPLgrid} and \textsc{fastNLO}.
Section~\ref{sec:multi-coupling-expansion} gives the details and, in particular, describes the problem of computing Monte Carlo weights for cross sections in a multi-coupling expansion; this section may be read first by readers unfamiliar with the programs mentioned previously.
In section~\ref{sec:perturbative-orders} we discuss a few properties of multi-coupling-expanded predictions.

@@

\item \textsc{PineAPPL} has been explicitly interfaced to \textsc{mg5\_aMC} (v3+)~\cite{Alwall:2014hca,Frederix:2018nkq},\footnote{A version of \textsc{mg5\_aMC} interfaced to \textsc{PineAPPL} can be downloaded from \url{https://launchpad.net/mg5amc-pineappl/trunk}, and it will be included as standard in the next \textsc{mg5\_aMC} release.} for which it supersedes \textsc{aMCfast}~\cite{Bertone:2014zva}. See appendix~\ref{app:sample-runcard} for an example of a runcard. While the usage of
\textsc{mg5\_aMC}(v3+) + \textsc{PineAPPL} is similar to that of 
\textsc{mg5\_aMC}(v2) + \textsc{PineAPPL} + \textsc{APPLgrid}, it is however more
efficient. For instance, the usage of the latter typically required a substantial amount of memory (close to \SI{120}{\giga\byte} for Drell--Yan), if a grid for a differential distribution with more than 10 bins was generated.
The memory usage is substantially reduced (roughly \SI{1}{\giga\byte}) after \enquote{optimisation} of the grids, i.e.\ an optimisation of the grid representation in memory.
However, this required a two-step procedure: first, to produce an unoptimised grid in order to identify those parts where the cross section is either zero or extremely suppressed; second, after those parts are removed, to fill an optimised grid with a small number of bins.
\textsc{PineAPPL} avoids this by using a more space-efficient representation from the start. This leads to substantially faster run times, in particular for simple processes, because the grids do not need to be optimised 
and their combination is faster.

@@

\paragraph{4-tuples}
A straightforward representation is given by 4-tuples, i.e.\ a list of the momentum fractions $x_1$ and $x_2$, the scale $Q^2$, and the phase-space weight $w$ for each phase-space point; 4-tuples are sufficient to reconstruct a total cross section.
For each combination $(a,b,k,l,m,n,o)$ (see section~\ref{sec:multi-coupling-expansion} for their definition) we save the following 4-tuples,
\begin{equation}
\left\{ x_1^i, x_2^i, Q^2_i, w^{(k,l,m,n,o)}_{ab} (x_1^i, x_2^i, Q^2_i, \mathcal{O}_i) \right\}_{i=1}^N \text{.} \label{eq:four-tuples}
\end{equation}
The reconstruction of the differential cross section is then done by simply multiplying the phase-space weights $w$ with PDFs evaluated with the correct arguments given in the 4-tuple and summing over all indices $a$, $b$, $k$, $l$, $m$, $n$, $o$, and $i$.

The 4-tuple representation is very easy to implement and test.
Furthermore, it reproduces the exact numerical value that is also calculated by the MC integrator.
However, the price one has to pay is the size of the 4-tuples.
For example, NLO QCD+EW Drell--Yan lepton-pair production (see section~\ref{sec:pineappl-example} and section~\ref{sec:dy-lepton-pair-production}) needs \SI{159}{\mega\byte} of storage for a target precision of \SI{1}{\percent} of the integrated cross section.
While this is an acceptable size, increasing the precision by an order of magnitude would require roughly 100 times the size, due to the Monte Carlo convergence that goes as $1/\sqrt{N}$ with $N$ being the number of 4-tuples.
With increasing size also the speed of the convolution degrades, because it basically becomes bound by the speed with which the 4-tuples can be read from disk.
However, due to the uncompressed nature of this representation it can serve as an intermediate format to develop and quickly cross check more space-efficient representations, which we will discuss next.

\paragraph{Lagrange-interpolation grid}
A different strategy is to partition a subset $H$ of the $(x_1, x_2, Q^2)$ space,
\begin{equation}
H = [x_\mathrm{min},x_\mathrm{max}]^2 \times [Q^2_\mathrm{min}, Q^2_\mathrm{max}] \ni (x_1, x_2, Q^2)
\end{equation}
along each axis into a small number of bins and to insert the phase-space weights $w$ into the corresponding discrete bin.
Using the bin centres and their values one already has a straightforward implementation of eq.~\eqref{eq:weight-map}, but given a finite number of bins usually yields an insufficient approximation for the cross section.
Increasing the number of bins improves the precision, but it also increases the space requirements.
Interpolation methods increase the precision using the same number of bins.

We use the same \enquote{Lagrange-interpolation grid} presented in ref.~\cite{Carli:2010rw} with the parameters published in ref.~\cite{Bertone:2014zva}, which give sufficient precision (see section~\ref{sec:results}).
For the sake of completeness, the following summarises the interpolation algorithm.

@@

To illustrate the filling step we give an example in figure~\ref{fig:grid}, where, for simplicity, we have chosen a static scale, so that we do not need to interpolate in the $\tau$ direction, and where we also limited the number of grid points to $N_y = 10$.
Each grid point has a numerical value $a_{i,j}$ associated, and the set of all numerical values $\{ a_{i,j} \}$ for all grid indices $(i,j) \in [0,N_y) \times [0,N_y)$ constitute \enquote{the grid}.
Inserting a specific weight $W = w (x_1, x_2) / \omega(x_1, x_2)$ into the grid is shown in figure~\ref{fig:grid} as a small black square, inside the larger grey one.
We have defined the grid points at specific positions, but the points given by the MC will land somewhere between them.
The interpolation order $s_y$ then defines a square with length $s_y + 1$ around its centre $(u(y(x_1)),u(y(x_2)))$, given by the MC.
All grid points with indices $(i,j)$ covered by the grey square are then updated according to the following formula:
\begin{equation}
a_{i,j} \leftarrow a_{i,j} + I_i(u(y(x_1))) I_j(u(y(x_2))) W(x_1,x_2) / \omega(x_1,x_2)
\label{eq:interpolation}
\end{equation}
with the Lagrange basis functions
\begin{equation}
I_i (u) = \prod_{\substack{k=k_i \\ k \neq i}}^{k_i + s_y} \frac{u-k}{i-k}
\end{equation}
where the product runs over all indices of the grid points covered by the grey square in figure~\ref{fig:grid}, starting from the smallest index in the square, $k_i$ and $k_j$.
Finally, we remapped
\begin{equation}
u(y) = \frac{y-y_\text{min}}{\Delta y} \text{,} \quad
\end{equation}
using $y_\text{min} = y(x_\text{max})$ and $y_\text{max} = y(x_\text{min})$ and the grid spacing $\Delta y = (y_\text{max} - y_\text{min})/(N_y-1)$, so that the integer part of $u(y)$ gives the grid index, e.g.\ $u(y_\text{min}) = 0$ and $u(y_\text{max}) = N_y - 1$, and the fractional part gives the location between the nearest grid points.

@@

In section~\ref{sec:multi-coupling-expansion} we labelled the different perturbative orders using the indices $k,l,m,n$.
Their values are process specific.
However, in general we define as leading order (LO) the set of contributions for all possible initial states $a b$ that lead to the same final state $X$, for which the sum of the coupling exponents in eq.~\eqref{eq:expansion} is smallest, i.e.\ $k + l = p$, where $p = \min (k+l)$.
This number is also process dependent and is usually determined by the number of external particles.
For many processes there is only one contribution at LO (in terms of $k$ and $l$), but this is not true in general. Indeed, when a process has multiple quark lines, colourless (photons, \dots) and coloured particles (gluons, \ldots) can be exchanged between them, making it possible to have more than one combinations of $\alphas$ and $\alpha$.
A typical example at the LHC is (on-shell) top-pair production, which has three different contributions at LO: $\mathcal{O} (\alpha^2)$, $\mathcal{O} (\alphas \alpha)$, and $\mathcal{O} (\alpha^2)$.
Each contribution receives a higher-order correction with an additional power of $\alphas$ or $\alpha$, which in general leads to at least two next-to-leading order (NLO) corrections.
The correction with the largest power in $\alphas$ can be unambiguously called \enquote{the} QCD correction, and the one with the largest power in $\alpha$ \enquote{the} EW correction.
All remaining corrections are of combined type, meaning that, in general, they cannot be attributed to either one of strong or electroweak origin. However, for the sake of simplicity and with a slight abuse of notation, it is customary to call
NLO QCD (EW) corrections those
corrections of order $\alphas$ ($\alpha$) times the couplings of the LO contribution 
with the largest power of $\alphas$. In the example above, NLO QCD and EW corrections to top pair production will denote
respectively those at $\mathcal{O} (\alphas^3 )$ and $\mathcal{O} (\alphas^2 \alpha)$. In this paper, in particular
when discussing results in sec.~\ref{sec:results}, we will explicitly list the orders that are considered in the cross
section at LO, NLO QCD and NLO QCD+EW accuracy.

>>>>

In this section we introduce \textsc{PineAPPL}.
We first describe the general scope and features of the library in comparison to \textsc{APPLgrid} and \textsc{fastNLO} in section~\ref{sec:library}.
Section~\ref{sec:multi-coupling-expansion} gives the details and, in particular, describes the problem of computing Monte Carlo weights for cross sections in a multi-coupling expansion; this section may be read first by readers unfamiliar with the programs mentioned previously.
In section~\ref{sec:perturbative-orders} we discuss a few properties of multi-coupling-expanded predictions.

@@

\item \textsc{PineAPPL} has been explicitly interfaced to \textsc{mg5\_aMC} (v3+)~\cite{Alwall:2014hca,Frederix:2018nkq},\footnote{A version of \textsc{mg5\_aMC} interfaced to \textsc{PineAPPL} can be downloaded from \url{https://launchpad.net/mg5amc-pineappl/trunk}, and it will be included as standard in the next \textsc{mg5\_aMC} release.} for which it supersedes \textsc{aMCfast}~\cite{Bertone:2014zva}; see appendix~\ref{app:sample-runcard} for an example of a runcard. While the usage of
\textsc{mg5\_aMC}(v3+) + \textsc{PineAPPL} is similar to that of 
\textsc{mg5\_aMC}(v2) + \textsc{aMCfast} + \textsc{APPLgrid}, it is however more
efficient. For instance, the usage of the latter typically required a substantial amount of memory (close to \SI{120}{\giga\byte} for Drell--Yan), if a grid for a differential distribution with more than 10 bins was generated.
The memory usage is substantially reduced (roughly \SI{1}{\giga\byte}) after \enquote{optimisation} of the grids, i.e.\ an optimisation of the grid representation in memory.
However, this required a two-step procedure: first, to produce an unoptimised grid in order to identify those parts where the cross section is either zero or extremely suppressed; second, after those parts are removed, to fill an optimised grid with a small number of bins.
\textsc{PineAPPL} avoids this by using a more space-efficient representation from the start. This leads to substantially faster run times, in particular for simple processes, because the grids do not need to be optimised 
and their combination is faster.

@@

\paragraph{4-tuples.}
A straightforward representation is given by 4-tuples, i.e.\ a list of the momentum fractions $x_1$ and $x_2$, the scale $Q^2$, and the phase-space weight $w$ for each phase-space point; 4-tuples are sufficient to reconstruct the differential cross sections.
For each combination $(a,b,k,l,m,n,o)$ (see section~\ref{sec:multi-coupling-expansion} for their definition) we save the following 4-tuples,
\begin{equation}
\left\{ x_1^i, x_2^i, Q^2_i, w^{(k,l,m,n,o)}_{ab} (x_1^i, x_2^i, Q^2_i, \mathcal{O}_i) \right\}_{i=1}^N \text{.} \label{eq:four-tuples}
\end{equation}
The reconstruction of the differential cross sections are then done by simply multiplying the phase-space weights $w$ with PDFs evaluated with the correct arguments given in the 4-tuple and summing over all indices $a$, $b$, $k$, $l$, $m$, $n$, $o$, and $i$.

The 4-tuple representation is very easy to implement and test.
Furthermore, it reproduces the exact numerical value that is also calculated by the MC integrator.
However, the price one has to pay is the size of the 4-tuples.
For example, NLO QCD+EW Drell--Yan lepton-pair production (see section~\ref{sec:pineappl-example} and section~\ref{sec:dy-lepton-pair-production}) needs \SI{159}{\mega\byte} of storage for a target precision of \SI{1}{\percent} of the integrated cross section.
While this is an acceptable size, increasing the precision by an order of magnitude would require roughly 100 times the size, due to the Monte Carlo convergence that goes as $1/\sqrt{N}$ with $N$ being the number of 4-tuples.
With increasing size also the speed of the convolution degrades, because it basically becomes bound by the speed with which the 4-tuples can be read from disk.
However, due to the uncompressed nature of this representation it can serve as an intermediate format to develop and quickly cross check more space-efficient representations, one of which we will discuss next.

\paragraph{Lagrange-interpolation grid.}
A different strategy is to partition a subset $H$ of the $(x_1, x_2, Q^2)$ space,
\begin{equation}
H = [x_\mathrm{min},x_\mathrm{max}]^2 \times [Q^2_\mathrm{min}, Q^2_\mathrm{max}] \ni (x_1, x_2, Q^2)
\end{equation}
along each axis into a small number of bins and to insert the phase-space weights $w$ into the corresponding discrete bin.
Using the bin centres and their values one already has a straightforward representation of eq.~\eqref{eq:weight-map}, but given a finite number of bins this approach usually yields an insufficient approximation for the cross section.
Increasing the number of bins improves the precision, but it also increases the space requirements.
This problem in turn in solved using interpolation methods, which increase the precision using the same number of bins.

We use the \enquote{Lagrange-interpolation grid} method presented in ref.~\cite{Carli:2010rw} with the parameters published in ref.~\cite{Bertone:2014zva}, which give sufficient precision (see section~\ref{sec:results}).
For the sake of completeness, the following summarises the interpolation algorithm.

@@

To illustrate the filling step we give an example in figure~\ref{fig:grid}, where, for simplicity, we have chosen a static scale, so that we do not need to interpolate in the $\tau$ direction, and where we also limited the number of grid points to $N_y = 10$.
Each grid point has a numerical value $a_{i,j}$ associated, and the set of all numerical values $\{ a_{i,j} \}$ for all grid indices $(i,j) \in [0,N_y) \times [0,N_y)$ constitute \enquote{the grid}.
Inserting a specific weight $W = w (x_1, x_2) / \omega(x_1, x_2)$ into the grid is shown in figure~\ref{fig:grid} as a small black square, inside the larger grey one.
We have defined the grid points at specific positions, but the points given by the MC will land somewhere between them.
The interpolation order $s_y$ then defines a square with length $s_y + 1$ around its centre $(u(y(x_1)),u(y(x_2)))$, given by the MC.
All grid points with indices $(i,j)$ covered by the grey square are then updated according to the following formula,
\begin{equation}
a_{i,j} \leftarrow a_{i,j} + I_i(u(y(x_1))) I_j(u(y(x_2))) W(x_1,x_2) / \omega(x_1,x_2) \text{,}
\label{eq:interpolation}
\end{equation}
with the Lagrange basis functions,
\begin{equation}
I_i (u) = \prod_{\substack{k=k_i \\ k \neq i}}^{k_i + s_y} \frac{u-k}{i-k} \text{,}
\end{equation}
where the product runs over all indices of the grid points covered by the grey square in figure~\ref{fig:grid}, starting from the smallest index in the square, $k_i$ and $k_j$.
Finally, we remap
\begin{equation}
u(y) = \frac{y-y_\text{min}}{\Delta y} \text{,} \quad
\end{equation}
using $y_\text{min} = y(x_\text{max})$ and $y_\text{max} = y(x_\text{min})$ and the grid spacing $\Delta y = (y_\text{max} - y_\text{min})/(N_y-1)$, so that the integer part of $u(y)$ gives the grid point index, e.g.\ $u(y_\text{min}) = 0$ and $u(y_\text{max}) = N_y - 1$, and the fractional part of $u(y)$ gives the relative location between the nearest grid points.

@@

In section~\ref{sec:multi-coupling-expansion} we labelled the different perturbative orders using the indices $k,l,m,n$; their values are process specific.
However, in general we define as leading order (LO) the set of contributions for all possible initial states $a b$ that lead to the same final state $X$, for which the sum of the coupling exponents in eq.~\eqref{eq:expansion} is smallest, i.e.\ $k + l = p$, where $p = \min (k+l)$.
This number is usually determined by the number of external particles.
For many processes there is only one contribution at LO (in terms of $k$ and $l$), but this is not true in general. Indeed, when a process has multiple quark lines, colourless (photons, \dots) and coloured particles (gluons, \ldots) can be exchanged between them, making it possible to have more than one combination of $\alphas$ and $\alpha$.
A typical example at the LHC is (on-shell) top-pair production, which has three different contributions at LO: $\mathcal{O} (\alpha^2)$, $\mathcal{O} (\alphas \alpha)$, and $\mathcal{O} (\alpha^2)$.
Each contribution receives a higher-order correction with an additional power of $\alphas$ or $\alpha$, which in general leads to at least two next-to-leading order (NLO) corrections.
The correction with the largest power in $\alphas$ can be unambiguously called \enquote{the} QCD correction, and the one with the largest power in $\alpha$ \enquote{the} EW correction.
All remaining corrections are of combined type, meaning that, in general, they cannot be attributed to either one of strong or electroweak origin. However, for the sake of simplicity and with a slight abuse of notation, it is customary to call
NLO QCD (EW) corrections those
corrections of order $\alphas$ ($\alpha$) times the couplings of the LO contribution 
with the largest power of $\alphas$. In the example above, NLO QCD and EW corrections to top pair production will denote
respectively those at $\mathcal{O} (\alphas^3 )$ and $\mathcal{O} (\alphas^2 \alpha)$. In this paper, in particular
when discussing results in sec.~\ref{sec:results}, we will explicitly list the orders that are considered in the cross
section at LO, NLO QCD and NLO QCD+EW accuracy.
